---
title: "Bill shape distinguishes Antarctic penguin species"
subtitle: "The Day-1 result — one source, a branded Typst PDF"
author:
  - name: Christophe Dervieux
    affiliation: Posit, PBC
date: last-modified
bibliography: references.bib
csl: apa.csl
citeproc: true
format:
  typst:
    papersize: a4
    margin:
      x: 2cm
      y: 2.5cm
    toc: true
    number-sections: true
    linestretch: 1.4
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(dplyr)
library(ggplot2)
library(ggrepel)
library(gt)
library(brand.yml)
library(prismatic)

brand <- read_brand_yml()
pal <- function(x) brand_color_pluck(brand, x)
# clr_darken returns an alpha-channel hex (#RRGGBBAA). Truncate to #RRGGBB.
tint <- function(x, f) substr(as.character(clr_darken(pal(x), f)), 1, 7)

brand_style <- function(t) {
  t |>
    theme_brand_gt(brand) |>
    opt_table_font(font = "Albert Sans") |>
    opt_row_striping() |>
    tab_style(
      style = list(cell_fill(color = pal("teal_lighter")),
                   cell_text(weight = "bold", color = pal("foreground"))),
      locations = cells_title(groups = "title")
    ) |>
    tab_style(
      style = list(cell_fill(color = pal("primary")),
                   cell_text(color = pal("background"), weight = "bold")),
      locations = cells_column_labels()
    ) |>
    tab_options(
      row.striping.background_color     = pal("teal_lighter"),
      table_body.hlines.color           = tint("teal_lighter", 0.12),
      table.border.top.color            = pal("primary"),
      table.border.top.width            = px(2),
      table.border.bottom.color         = pal("primary"),
      table.border.bottom.width         = px(2),
      column_labels.border.bottom.color = pal("primary"),
      column_labels.border.bottom.width = px(2),
      heading.align                     = "left"
    )
}

data(penguins)
penguins <- penguins |> filter(!is.na(bill_len), !is.na(bill_dep), !is.na(body_mass))
```

# Introduction

The base-R `penguins` dataset records body measurements for `r nrow(penguins)` penguins of three
species collected at Palmer Station, Antarctica [@gorman2014]. It reached the R community first
through the **palmerpenguins** package [@horst2020]. Here we ask a simple question: **does bill
shape tell the species apart?** We render the answer as a branded PDF with **Typst**, a modern
typesetting system bundled with Quarto that does not require LaTeX.

# Species at a glance

```{r}
#| label: tbl-species
summary_tbl <- penguins |>
  summarise(
    n = n(),
    bill_len = mean(bill_len),
    bill_dep = mean(bill_dep),
    body_mass = mean(body_mass),
    .by = species
  )

summary_tbl |>
  gt() |>
  tab_header(
    title = "Penguins of the Palmer Archipelago",
    subtitle = "Mean measurements per species"
  ) |>
  tab_spanner(label = md("**Bill (mm)**"), columns = c(bill_len, bill_dep)) |>
  cols_label(
    species = "Species", n = "N",
    bill_len = "Length", bill_dep = "Depth", body_mass = "Body mass (g)"
  ) |>
  fmt_number(columns = c(bill_len, bill_dep), decimals = 1) |>
  fmt_number(columns = body_mass, decimals = 0, sep_mark = " ") |>
  brand_style()
```

Of the three species, `r summary_tbl$species[which.max(summary_tbl$body_mass)]` has the
`#highlight(fill: rgb("#D1E5E6"))[highest mean body mass]`{=typst}
(`r format(round(max(summary_tbl$body_mass)), big.mark = " ")` g).

# Bill length versus depth

```{r}
#| label: fig-bill
#| fig-cap: "Bill length versus depth by species. The single heaviest penguin is labeled."
#| fig-alt: >-
#|   Scatter plot of bill depth against bill length for penguins of three species, each drawn in a
#|   distinct teal shade and point shape. The three species form separate clusters. The single
#|   heaviest penguin in the dataset is highlighted and labeled with its species.
heaviest <- penguins |> slice_max(body_mass, n = 1)

ggplot(penguins, aes(x = bill_len, y = bill_dep, color = species, shape = species)) +
  geom_point(alpha = 0.9, size = 2) +
  geom_text_repel(
    data = heaviest,
    aes(label = paste0(species, " · ", body_mass, " g")),
    color = pal("foreground"), fontface = "bold",
    size = 3, min.segment.length = 0, show.legend = FALSE
  ) +
  scale_color_manual(values = c(
    "Adelie"    = pal("code_blue"),   # dark
    "Chinstrap" = pal("primary"),     # mid teal
    "Gentoo"    = pal("teal_light")   # light teal
  )) +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)",
       color = "Species", shape = "Species") +
  theme_minimal(base_size = 12) +
  theme_brand_ggplot2(brand)
```

The three species separate cleanly in bill shape. This is the same figure from the HTML report, now
shown in the branded PDF.

# Conclusion

One `.qmd`, one `_brand.yml`, and `format: typst` give a typeset document with citations. They apply
the RaukR typography and palette to the document, tables, and figures.

## References {.unnumbered}

::: {#refs}
:::
