---
title: "data.table"
author: "Markus Mayrhofer"
description: "Fast, memory-efficient tabular data wrangling in R, compared with base R and the tidyverse."
format: html
---

::: {.callout-note}
## Coding exercise

We recommend that you run the code sections as you work through this exercise.
You can copy-paste into a Quarto or R script and run line by line using
{{< kbd mac=Command-Enter win=Ctrl-Enter linux=Ctrl-Enter >}}.
:::

```{r}
#| label: setup
#| message: false
library(data.table)   # the topic of this lab
library(dplyr)        # tidyverse reference implementations
library(readr)        # read_tsv(), for the file-reading benchmark
library(bench)        # benchmarking
library(ggplot2)      # benchmark plots
theme_set(theme_minimal())
setDTthreads(percent = 100)   # use all cores; check with getDTthreads()
```

## Introduction

`data.table` is an R package for working with **rectangular / tabular data**: rows
are observations, columns are variables, each cell is one value. Most
data-wrangling tasks are combinations of four basic actions:

| Question | Operation |
|---|---|
| Which rows do I keep? | **filter** rows |
| Which columns do I keep or compute? | **select / mutate** columns |
| Should I calculate separately for each group? | **group by** |
| Do I need information from another table? | **join** |

**Where it comes from.** `data.table` was created by Matt Dowle, with version 1.0
on CRAN in April 2006, and was long co-developed with Arun Srinivasan. It is now
maintained by a wider team (Tyson Barrett is the current CRAN maintainer, with
Jan Gorecki, Michael Chirico, Toby Hocking and others). It grew out of a practical
problem: row-wise operations on a base R `data.frame` -- filtering, `rbind`,
`merge` -- rebuild every column, which becomes slow and memory-hungry once tables
get large. `data.table` is written largely in C, is
multi-threaded, adds or changes columns *by reference* (without copying), and ships
with fast file readers (`fread` / `fwrite`). It keeps a deliberately small
dependency footprint -- just base R -- which is part of why it stays fast and
stable. The cost is a terse, bracket-based syntax, which this lab unpacks.

In this lab we express each of these in **three styles** and compare them:

- **base R**: the tools that ship with R (`[`, `$`, `subset()`, `tapply()`, `aggregate()`, `merge()`).
- **tidyverse** (`dplyr`): a sequence of named verbs joined by the pipe `|>`. Shown here only for comparison; many of you will already know it.
- **data.table**: the focus of today. Everything happens inside `DT[i, j, by]`.

::: {.callout-note}
## The one idea to remember: `DT[i, j, by]`

```
DT[ i ,  j ,  by ]
|  |    |    `-- group by these columns
|  |    `------- compute, select, or update columns
|  `------------ keep or match these rows
`--------------- the data.table object
```

Read it as: **take rows `i`, do `j`, separately within each `by` group.**
If there is no grouping, leave `by` out. If there is no row filter, leave `i`
empty but keep the comma: `DT[, j]`.

Two shorthands appear constantly:

- `.N` = number of rows in the current group (like `dplyr::n()`).
- `.()` = shorthand for `list()`; used in `j` to return or name columns.
:::

## Data: a simulated genotyping cohort

We simulate genetic data because it produces **large tables**, which is
exactly where `data.table` earns its keep. No files to download: the generator
below is fully reproducible. The biological terms are only labels; the same
row/column/group/join ideas apply to any large table.

| Term | Meaning here |
|---|---|
| sample | one person or specimen |
| variant | one genomic position where samples may differ |
| genotype | number of alternate copies: 0, 1, or 2; **-1 means missing** |
| read depth | how many sequencing reads cover a site |
| genotype quality | confidence score for the genotype call |
| allele frequency | average alternate-copy frequency across samples |

The data has three linked tables:

- **`samples`**: one row per sample (population, sex, batch, case/control).
- **`variants`**: one row per variant (chromosome, position, gene, consequence, frequency).
- **`genotypes`**: one row per sample x variant pair (genotype, read depth, quality). This is the **tall** table.

```{r}
#| label: simulate
#| code-fold: true
#| code-summary: "Run to generate the data -- click to expand"
simulate_genotyping <- function(n_samples = 500, n_variants = 4000, seed = 42) {
  set.seed(seed)

  # samples: one row per sample
  pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")
  samples <- data.table(
    sample_id  = sprintf("S%05d", seq_len(n_samples)),
    population = sample(pops, n_samples, replace = TRUE,
                        prob = c(0.22, 0.14, 0.20, 0.30, 0.14)),
    sex        = sample(c("M", "F"), n_samples, replace = TRUE),
    batch      = sample(sprintf("batch%02d", 1:8), n_samples, replace = TRUE),
    phenotype  = rbinom(n_samples, 1, 0.35)   # 1 = case, 0 = control
  )

  # variants: one row per variant
  chroms <- c(1:22, "X")
  conseq <- c("synonymous", "missense", "stop_gained", "splice_region", "intron", "UTR")
  variants <- data.table(
    variant_id  = sprintf("rs%07d", seq_len(n_variants)),
    chrom       = sample(chroms, n_variants, replace = TRUE),
    pos         = sample(1e6:2.5e8, n_variants, replace = TRUE),
    ref         = sample(c("A", "C", "G", "T"), n_variants, replace = TRUE),
    alt         = sample(c("A", "C", "G", "T"), n_variants, replace = TRUE),
    gene        = sprintf("GENE%04d", sample(1:1500, n_variants, replace = TRUE)),
    consequence = sample(conseq, n_variants, replace = TRUE,
                         prob = c(0.30, 0.25, 0.03, 0.07, 0.25, 0.10)),
    maf         = round(rbeta(n_variants, 0.6, 6), 4)   # skewed toward rare
  )
  flip <- variants$alt == variants$ref
  variants[flip, alt := fifelse(ref == "A", "G", "A")]

  # genotypes: one row per sample x variant (the tall table)
  geno <- CJ(sample_id = samples$sample_id, variant_id = variants$variant_id)
  geno[variants, maf := i.maf, on = "variant_id"]
  n <- nrow(geno)
  p <- geno$maf
  u <- runif(n)
  geno[, genotype := fifelse(u < (1 - p)^2, 0L,
                       fifelse(u < (1 - p)^2 + 2 * p * (1 - p), 1L, 2L))]
  geno[runif(n) < 0.02, genotype := -1L]                # 2% missing
  geno[, read_depth       := rpois(n, lambda = 30)]
  geno[, genotype_quality := pmin(99L, rpois(n, lambda = 45))]
  geno[, maf := NULL]
  setcolorder(geno, c("sample_id", "variant_id", "genotype", "read_depth", "genotype_quality"))

  list(genotypes = geno[], variants = variants[], samples = samples[])
}

dat       <- simulate_genotyping(n_samples = 500, n_variants = 4000, seed = 42)
genotypes <- dat$genotypes
variants  <- dat$variants
samples   <- dat$samples

# Plain data.frame copies. The base R *and* tidyverse code below both use these,
# so each style is measured on its own native input: base R indexing on a
# data.frame, dplyr verbs on a data.frame, and data.table's `[` on a data.table.
genotypes_df <- as.data.frame(genotypes)
variants_df  <- as.data.frame(variants)
samples_df   <- as.data.frame(samples)

dim(genotypes)                                  # 500 x 4000 = 2,000,000 rows
format(object.size(genotypes), units = "MB")
head(genotypes, 4)
```

::: {.callout-tip}
Exactly 2,000,000 rows (500 samples x 4000 variants), the same on every machine
because the seed is fixed. The size is intentional: small examples hide the
memory and copying behaviour that matters in real analyses. Only the benchmark
*timings* later will differ from one computer to another; the data and the
results will not.
:::

## Core operations

Each operation below is shown in all three styles. Watch where the row filter,
the column calculation, and the grouping live in each.

::: {.callout-note}
## Why the tidyverse code uses `genotypes_df`
`data.table` inherits from `data.frame`, so dplyr verbs *do* work on a
`data.table` -- but dplyr then has to convert it, which makes dplyr look slower
and more memory-hungry than it really is. To keep the later benchmarks fair, the
base R and tidyverse code both work on the plain `data.frame` copies
(`genotypes_df`, `variants_df`, `samples_df`) while the `data.table` code works
on the `data.table`. Each style is measured on its own native input.

A `tibble` would work equally well: it is a `data.frame` underneath and benchmarks
within a few percent. Plain `data.frame` also keeps the printed results in the same
style as the other two engines.
:::

### Filter rows

Get genotype rows where the genotype is a homozygous alternate call and read
depth is adequate (showing top three).

```{r}
#| label: op-filter
# base R
head(genotypes_df[genotypes_df$genotype == 2 & genotypes_df$read_depth >= 20, ], 3)

# tidyverse
genotypes_df |> filter(genotype == 2, read_depth >= 20) |> head(3)

# data.table: the row condition goes in i, before the first comma
genotypes[genotype == 2 & read_depth >= 20][1:3]
```

Under the hood: all three build a TRUE/FALSE vector over the rows and keep the
`TRUE` ones. All three return a **new** table -- the original is untouched -- so you
pay memory for the rows you keep, and you would normally assign the result a name.

### Select columns

```{r}
#| label: op-select
# base R
head(variants_df[, c("variant_id", "gene", "maf")], 3)

# tidyverse
variants_df |> select(variant_id, gene, maf) |> head(3)

# data.table: selected columns go in j as .(...)
variants[, .(variant_id, gene, maf)][1:3]
```

Under the hood: all three return a **new** table object, but the data vectors
themselves are not duplicated until something modifies them.

### Add a column

```{r}
#| label: op-mutate
# base R (shallow copy: the other columns are shared with variants_df)
v_tmp <- variants_df
v_tmp$is_rare <- v_tmp$maf < 0.01
head(v_tmp[, c("variant_id", "maf", "is_rare")], 3)

# tidyverse (returns a new object)
variants_df |> mutate(is_rare = maf < 0.01) |>
  select(variant_id, maf, is_rare) |> head(3)

# data.table: := adds/updates a column by reference (no copy)
variants[, is_rare := maf < 0.01]
variants[, .(variant_id, maf, is_rare)][1:3]
```

::: {.callout-warning}
## Reference semantics
`:=` changes `variants` **in place**: there is no new object to assign, and every
other reference to that table sees the change. Use `copy(variants)` first if you
need an untouched snapshot. A function that receives a `data.table` and uses `:=`
can modify the caller's object unless it copies internally.

The base R and tidyverse versions above are not expensive either: a `data.frame`
is a list of columns, so they share the unchanged columns and duplicate only the
column you touch. The copies that actually cost you are row-wise -- filters,
`rbind`, joins -- which have to rebuild every column.
:::

### Group and summarise

For each sample, among called genotypes, compute the mean read depth and the
number of called genotypes.

```{r}
#| label: op-group
# base R (aggregate() returns a matrix column, so flatten and rename)
called <- genotypes_df[genotypes_df$genotype >= 0, ]
agg <- aggregate(read_depth ~ sample_id, data = called,
                 FUN = function(x) c(mean_depth = mean(x), n_called = length(x)))
agg <- do.call(data.frame, agg)
names(agg) <- c("sample_id", "mean_depth", "n_called")
head(agg, 3)

# tidyverse
genotypes_df |>
  filter(genotype >= 0) |>
  group_by(sample_id) |>
  summarise(mean_depth = mean(read_depth), n_called = n()) |>
  head(3)

# data.table: i filters, j computes, by defines the groups
genotypes[genotype >= 0,
          .(mean_depth = mean(read_depth), n_called = .N),
          by = sample_id][1:3]
```

### Join tables

Attach each variant's `consequence` and `gene` to the genotype rows, using
`variant_id` as the shared key.

```{r}
#| label: op-join
# base R (merge() reorders around the join key; re-sort to match the others)
m <- merge(genotypes_df, variants_df[, c("variant_id", "consequence", "gene")],
           by = "variant_id")
m <- m[order(m$sample_id, m$variant_id),
       c("sample_id", "variant_id", "genotype", "consequence", "gene")]
head(m, 3)

# tidyverse
genotypes_df |>
  inner_join(select(variants_df, variant_id, consequence, gene), by = "variant_id") |>
  select(sample_id, variant_id, genotype, consequence, gene) |>
  head(3)

# data.table: x[y, on = "key"]
variants[genotypes, on = "variant_id",
         .(sample_id, variant_id, genotype, consequence, gene)][1:3]
```

## Keys

The joins above used `on =` to match columns ad-hoc. If you look up or join on the
same column repeatedly, it is worth setting a **key**.

A key is one or more columns that a `data.table` is physically **sorted by**.
`setkey(DT, col)` reorders the rows of `DT` by `col` -- in place, by reference --
and records that order.

```{r}
#| label: keys
setkey(genotypes, sample_id)
key(genotypes)
```

Because the rows are now sorted by `sample_id`, `data.table` can find rows by
**binary search** rather than scanning the whole table:

```{r}
#| label: keys-lookup
genotypes["S00001"][1:3]     # keyed lookup: one sample's rows, found directly
```

A key speeds up three things on the key column(s):

- **Row lookups**, as above.
- **Repeated joins**: two tables keyed on the same column join by a fast sorted merge.
- **Grouping** by the key column.

You do not always need a key -- the `on =` joins above work without one -- but set a
key when you will hit the same column many times, so the sort cost is paid once.
`setindex()` gives a similar speed-up without physically re-ordering the table.

## Benchmarks: time and memory

`bench::mark()` runs each expression and records **elapsed time**, **memory
allocated**, and **garbage collections**. Time matters because slow code interrupts
analysis. Memory matters because large tables can fail when R runs out of RAM.
`check = FALSE` allows small differences in output shape between engines.

```{r}
#| label: bench-fig-helper
#| code-fold: true
#| code-summary: "Helper: plot a bench::mark result (time, and memory when measured)"
bench_fig <- function(b, engines) {
  cols <- c("#D55E00", "#0072B2", "#009E73")   # base_R, middle engine, data.table
  names(cols) <- engines

  long <- data.frame(engine = as.character(b$expression),
                     metric = "median time (ms)",
                     value  = as.numeric(b$median) * 1000)
  if (!is.null(b$mem_alloc) && !all(is.na(b$mem_alloc))) {
    long <- rbind(long, data.frame(
      engine = as.character(b$expression),
      metric = "memory allocated (MB)",
      value  = as.numeric(b$mem_alloc) / 1024^2))
  }
  # rev() so the first engine (base_R) sits on top after coord_flip
  long$engine <- factor(long$engine, levels = rev(engines))
  long$metric <- factor(long$metric,
                        levels = c("median time (ms)", "memory allocated (MB)"))

  # horizontal bars without coord_flip, so facet free scales work per panel
  ggplot(long, aes(x = value, y = engine, fill = engine)) +
    geom_col(width = 0.65, show.legend = FALSE, orientation = "y") +
    geom_text(aes(label = round(value, 1)), hjust = -0.15, size = 3.3) +
    facet_wrap(~ metric, scales = "free_x") +   # each metric gets its own value axis
    scale_fill_manual(values = cols) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
    labs(x = NULL, y = NULL) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.major.y = element_blank())
}
```

### Grouped aggregation: per-variant allele frequency

For each variant, among called genotypes, allele frequency is
`sum(genotype) / (2 * n_called)`.

```{r}
#| label: bench-agg
bench_agg <- bench::mark(
  base_R = {
    ca <- genotypes_df[genotypes_df$genotype >= 0, c("variant_id", "genotype")]
    tapply(ca$genotype, ca$variant_id, sum) /
      (2 * tapply(ca$genotype, ca$variant_id, length))
  },
  tidyverse = genotypes_df |>
    filter(genotype >= 0) |>
    group_by(variant_id) |>
    summarise(af = sum(genotype) / (2 * n()), .groups = "drop"),
  data.table = genotypes[genotype >= 0,
                         .(af = sum(genotype) / (2 * .N)),
                         by = variant_id],
  check = FALSE,
  iterations = 3
)
bench_agg[, c("expression", "median", "mem_alloc", "n_gc")]
```

```{r}
#| label: fig-bench-agg
#| fig-cap: "Grouped aggregation over 2,000,000 genotype rows: median run time and memory allocated, per engine. Lower is better."
#| fig-height: 3
bench_fig(bench_agg, c("base_R", "tidyverse", "data.table"))
```

::: {.callout-note}
## Reading this result
On this run `dplyr` is the **fastest** on time: a
grouped `summarise()` over a few thousand groups is well optimised, so
`data.table` has little headroom left on time for this particular task.

Where `data.table` does win here is **memory**: it allocates several times less
than either alternative, because it groups without materialising large
intermediate copies. If your table is big enough that RAM is the binding
constraint, that difference decides whether the job runs at all.

`data.table`'s *time* advantage shows up in the next two benchmarks, the join and the file
read. This is why we measure instead of asserting: which tool wins depends
on the operation, the shape of the data, and whether time or memory is scarce.
Run this on your own machine and the results may differ slightly.
:::

### Three-table join

Attach variant `consequence` + `gene` and sample `population` to every genotype row.

::: {.callout-warning}
## Read this benchmark with one caveat
The `setkey()` calls happen **before** `bench::mark()`, so the one-time cost of
sorting the tables is **not** included in the `data.table` timing, while base R
and dplyr pay their full cost inside the timed expression. That matches a real
application -- you sort once, then join and look up many times -- but this benchmark
performs only a *single* join. For a genuine one-off join you would have to add the 
sort cost to `data.table`'s time, reducing its advantage slightly.
:::

```{r}
#| label: bench-join
# keyed joins are sorted-merge joins (fast); set keys once, outside the timing
setkey(genotypes, variant_id)
setkey(variants,  variant_id)

bench_join <- bench::mark(
  base_R = {
    m1 <- merge(genotypes_df, variants_df[, c("variant_id", "consequence", "gene")],
                by = "variant_id")
    merge(m1, samples_df[, c("sample_id", "population")], by = "sample_id")
  },
  tidyverse = genotypes_df |>
    inner_join(select(variants_df, variant_id, consequence, gene), by = "variant_id") |>
    inner_join(select(samples_df,  sample_id, population),          by = "sample_id"),
  data.table = samples[
    variants[genotypes, on = "variant_id",
             .(sample_id, variant_id, genotype, consequence, gene)],
    on = "sample_id"],
  check = FALSE,
  iterations = 2
)
bench_join[, c("expression", "median", "mem_alloc")]
```

```{r}
#| label: fig-bench-join
#| fig-cap: "Three-table join over 2,000,000 genotype rows: median run time and memory allocated, per engine. Lower is better."
#| fig-height: 3
bench_fig(bench_join, c("base_R", "tidyverse", "data.table"))
```

### Reading a file

Large analyses often start by reading a file. Write the genotype table to a
temporary TSV, then compare read speeds.

```{r}
#| label: bench-io
tmp <- tempfile(fileext = ".tsv")
data.table::fwrite(genotypes, tmp, sep = "\t")
cat("file size:", round(file.info(tmp)$size / 1e6, 1), "MB\n")

bench_io <- bench::mark(
  base_R     = read.delim(tmp),
  readr      = readr::read_tsv(tmp, show_col_types = FALSE) |> as.data.frame(),
  data.table = data.table::fread(tmp),
  check = FALSE, iterations = 3, memory = FALSE
)
bench_io[, c("expression", "median")]
unlink(tmp)
```

```{r}
#| label: fig-bench-io
#| fig-cap: "Reading a ~50 MB TSV: median run time, per engine. Lower is better. (Memory is not profiled for file readers, so time only.)"
#| fig-height: 2.4
bench_fig(bench_io, c("base_R", "readr", "data.table"))
```

## Exercises

Each exercise gives the base R and tidyverse answers so you know the target
output. Write the `data.table` version yourself, then open the folded solution to
check.

### Per-sample call rate

For each sample, what fraction of genotypes are non-missing (`genotype >= 0`)?
Show the 3 samples with the **lowest** call rate.

```{r}
#| label: ex1-ref
# base R
cr <- sort(tapply(genotypes_df$genotype >= 0, genotypes_df$sample_id, mean))
head(data.frame(sample_id = names(cr), call_rate = as.numeric(cr)), 3)

# tidyverse
genotypes_df |>
  group_by(sample_id) |>
  summarise(call_rate = mean(genotype >= 0)) |>
  arrange(call_rate) |>
  head(3)
```

```{r}
#| label: ex1-sol
#| code-fold: true
#| code-summary: "Show data.table solution"
genotypes[, .(call_rate = mean(genotype >= 0)), by = sample_id][order(call_rate)][1:3]
```

### High-quality missense rows per gene

Among genotype rows with `genotype_quality >= 30` and `genotype > 0`, joined to
variant annotation and keeping `consequence == "missense"`, which 3 genes have the
most such rows?

```{r}
#| label: ex2-ref
# base R
g  <- genotypes_df[genotypes_df$genotype_quality >= 30 & genotypes_df$genotype > 0, ]
mg <- merge(g, variants_df[, c("variant_id", "consequence", "gene")], by = "variant_id")
tab <- sort(table(mg$gene[mg$consequence == "missense"]), decreasing = TRUE)
head(data.frame(gene = names(tab), n = as.integer(tab)), 3)

# tidyverse
genotypes_df |>
  filter(genotype_quality >= 30, genotype > 0) |>
  inner_join(variants_df, by = "variant_id") |>
  filter(consequence == "missense") |>
  count(gene, sort = TRUE) |>
  head(3)
```

```{r}
#| label: ex2-sol
#| code-fold: true
#| code-summary: "Show data.table solution"
variants[genotypes[genotype_quality >= 30 & genotype > 0], on = "variant_id"][
  consequence == "missense", .N, by = gene][order(-N)][1:3]
```

### Kept calls per population

Joining genotypes to `samples`, for each `population` how many genotype rows pass
`read_depth >= 15 & genotype_quality >= 25`?

```{r}
#| label: ex3-ref
# base R
mp   <- merge(genotypes_df, samples_df[, c("sample_id", "population")], by = "sample_id")
keep <- mp$read_depth >= 15 & mp$genotype_quality >= 25
kept <- tapply(keep, mp$population, sum)
data.frame(population = names(kept), kept = as.integer(kept))

# tidyverse
genotypes_df |>
  inner_join(select(samples_df, sample_id, population), by = "sample_id") |>
  group_by(population) |>
  summarise(kept = sum(read_depth >= 15 & genotype_quality >= 25))
```

```{r}
#| label: ex3-sol
#| code-fold: true
#| code-summary: "Show data.table solution"
samples[genotypes, on = "sample_id"][
  , .(kept = sum(read_depth >= 15 & genotype_quality >= 25)), by = population][order(population)]
```

## Cheat sheet

| Task | base R | tidyverse | data.table |
|---|---|---|---|
| filter rows | `x[x$a > 1, ]` | `filter(x, a > 1)` | `x[a > 1]` |
| select cols | `x[, c("a", "b")]` | `select(x, a, b)` | `x[, .(a, b)]` |
| new col | `x$z <- x$a + x$b` | `mutate(x, z = a + b)` | `x[, z := a + b]` |
| summarise | `aggregate(a ~ g, x, mean)` | `summarise(x, m = mean(a))` | `x[, .(m = mean(a))]` |
| group | (via `aggregate` / `tapply`) | `group_by(x, g)` | `x[, ..., by = g]` |
| count | `table(x$g)` | `count(x, g)` | `x[, .N, by = g]` |
| arrange | `x[order(-x$a), ]` | `arrange(x, desc(a))` | `x[order(-a)]` |
| join | `merge(x, y, by = "k")` | `inner_join(x, y, by = "k")` | `x[y, on = "k"]`, or `merge(x, y, by = "k")` |
| read tsv | `read.delim(f)` | `read_tsv(f)` | `fread(f)` |
| write tsv | `write.table(x, f)` | `write_tsv(x, f)` | `fwrite(x, f)` |

## A caveat: whose method is actually running?

The "base R" column above assumes a plain `data.frame`. That matters, because
`data.table` inherits from `data.frame` **and** supplies its own methods for
several base R generics -- so identical-looking code can dispatch to a completely
different implementation depending on what you hand it. `merge()` is the clearest
example:

```{r}
#| label: merge-dispatch
dt_x <- data.table(k = 1:3, v = 1:3)
df_x <- as.data.frame(dt_x)

"merge.data.table" %in% as.character(methods("merge"))  # data.table defines one
class(merge(dt_x, dt_x, by = "k"))                      # -> merge.data.table ran
class(merge(df_x, df_x, by = "k"))                      # -> merge.data.frame ran
```

Hand `merge()` two `data.table`s and you get data.table's own C join, which on
millions of rows is more than an order of magnitude faster than base R's
`merge.data.frame` on the same data -- so `merge()` is a perfectly good
`data.table` join, not just a base R one. The same applies to `[`: a `data.table`
uses `[.data.table`, not `[.data.frame`.

This is exactly why the base R benchmarks above use the `_df` copies: to measure
base R itself, rather than `data.table` wearing a base R coat. Functions with no
`data.table` method -- `aggregate()`, `tapply()`, `table()`, `order()` -- are
genuinely base R whatever you pass them.

## Where to go next

- `vignette("datatable-intro")` and the other built-in `data.table` vignettes
- `?data.table` for the full `[i, j, by]` reference
- The `dtplyr` package: write dplyr code, get data.table speed
- `setDTthreads()` / `getDTthreads()` to tune parallelism

## Session

<details>
  <summary>Session info</summary>

```{r}
#| label: session
sessionInfo()
```

</details>
