---
title: "Interactive Visualization in R"
author: "Roy Francis"
description: "Hands-on exercises with plotly and highcharter."
format: html
---

::: {.callout-note}
This lab focuses on practical interactive charting with `plotly` and `highcharter`.
You will build and refine charts step-by-step using the `iris` and `mtcars` datasets.
:::

## Learning Objectives

- Build interactive scatter, line, and bar charts with `plotly`.
- Build interactive charts with `highcharter` using both quick and explicit APIs.
- Customize tooltips, legends, axes, themes, and interactions.
- Combine charts, expose controls, and choose an appropriate chart type.

## Setup

Install packages if not already installed.

```{r}
#| eval: false
install.packages(c("plotly", "highcharter", "dplyr", "ggplot2"))
```

Load packages.

```{r}
library(dplyr)
library(ggplot2)
library(plotly)
library(highcharter)
```

Load datasets.

```{r}
data("iris")
data("mtcars")
mtcars$model <- rownames(mtcars)
```

## **plotly** 

### Basics

Create an interactive scatter plot of `Sepal.Length` vs `Sepal.Width`, colored by `Species`. Refer to the [plotly R reference](https://plotly.com/r/) for information.

```{r}
plot_ly(
  data = iris,
  x = ~Sepal.Length,
  y = ~Sepal.Width,
  color = ~Species,
  type = "scatter",
  mode = "markers"
)
```

Add custom hover text that shows species and petal dimensions.


::: {.callout-tip}

HTML tag `<br>` is used to create line breaks in the hover text.

:::

```{r}
plot_ly(
  data = iris,
  x = ~Sepal.Length,
  y = ~Sepal.Width,
  color = ~Species,
  text = ~paste(
    "Species:", Species,
    "<br>Petal Length:", Petal.Length,
    "<br>Petal Width:", Petal.Width
  ),
  hoverinfo = "text",
  type = "scatter",
  mode = "markers"
)
```

### Customization

Add a title, axis labels, and place the legend horizontally.

```{r}
plot_ly(
  data = iris,
  x = ~Sepal.Length,
  y = ~Sepal.Width,
  color = ~Species,
  type = "scatter",
  mode = "markers"
) |>
  layout(
    title = "Iris Sepal Dimensions",
    xaxis = list(title = "Sepal Length"),
    yaxis = list(title = "Sepal Width"),
    legend = list(orientation = "h")
  )
```

### ggplotly

Convert a ggplot to plotly with `ggplotly()`.

```{r}
p <- ggplot(iris, aes(Petal.Length, Petal.Width, color = Species)) +
  geom_point()

ggplotly(p)
```

### Mapping variables and hover templates

Map multiple variables to marker appearance. Here, point color shows the number
of cylinders and marker size shows horsepower. The `customdata` argument keeps
the model name available for the hover template without displaying it as a
separate aesthetic.

```{r}
plot_ly(
  data = mtcars,
  x = ~wt,
  y = ~mpg,
  color = ~factor(cyl),
  size = ~hp,
  sizes = c(8, 32),
  customdata = ~model,
  type = "scatter",
  mode = "markers",
  marker = list(opacity = 0.75)
) |>
  layout(
    title = "Fuel economy, weight, and engine size",
    xaxis = list(title = "Weight (1000 lb)"),
    yaxis = list(title = "Miles per gallon"),
    legend = list(title = list(text = "Cylinders"))
  ) |>
  style(
    hovertemplate = paste0(
      "<b>%{customdata}</b><br>",
      "Weight: %{x:.2f}<br>",
      "MPG: %{y:.1f}<br>",
      "Horsepower: %{marker.size}<extra></extra>"
    )
  )
```

### Time-series charts

Use `add_lines()` to make a time-series-style chart. `economics` is included
with ggplot2 and gives us a date variable to work with.

```{r}
economics_small <- economics |>
  filter(date >= as.Date("2005-01-01")) |>
  transmute(date, unemployment = unemploy / 1000, population = pop / 1000)

plot_ly(economics_small, x = ~date) |>
  add_lines(y = ~unemployment, name = "Unemployed (millions)") |>
  layout(
    title = "US unemployment after 2005",
    xaxis = list(title = "Date", rangeslider = list(visible = TRUE)),
    yaxis = list(title = "People (millions)")
  )
```

### Subplots

Combine two related iris scatter plots side-by-side with `subplot()`. Both use the species color mapping, so one legend describes the points in both
panels. `highlight_key()` assigns each flower a unique identifier.

::: {.column-body-outset}

```{r}
iris_keyed <- iris |>
  mutate(id = row_number()) |>
  highlight_key(~id)

sepal_plot <- plot_ly(
  iris_keyed,
  x = ~Sepal.Length,
  y = ~Sepal.Width,
  color = ~Species,
  type = "scatter",
  mode = "markers"
) |>
  layout(
    xaxis = list(title = "Sepal length"),
    yaxis = list(title = "Sepal width")
  )

petal_plot <- plot_ly(
  iris_keyed,
  x = ~Petal.Length,
  y = ~Petal.Width,
  color = ~Species,
  type = "scatter",
  mode = "markers",
  showlegend = FALSE
) |>
  layout(
    xaxis = list(title = "Petal length"),
    yaxis = list(title = "Petal width")
  )

subplot(
  sepal_plot, petal_plot,
  nrows = 1,
  titleX = TRUE,
  titleY = TRUE
) |>
  layout(title = "Iris sepal and petal dimensions")
```

:::

### Linked subplots

Continue with the same plot objects and add `highlight()`. With `dragmode =
"select"`, draw a box around points in either panel to brush the same flowers
in the other panel. Double-click an empty area to clear the selection.

::: {.column-body-outset}

```{r}
subplot(sepal_plot, petal_plot, nrows = 1, titleX = TRUE, titleY = TRUE) |>
  layout(
    title = "Brush flowers across sepal and petal measurements",
    dragmode = "select"
  ) |>
  highlight(
    on = "plotly_selected",
    off = "plotly_doubleclick",
    dynamic = TRUE
  )
```

:::

### Dropdown controls

Add a dropdown control with `updatemenus`. This lets the reader switch between
two y variables without recreating the chart.

```{r}
plot_ly(
  mtcars, 
  x = ~wt, 
  y = ~mpg, 
  type = "scatter", 
  mode = "markers"
) |>
  layout(
    title = "Choose a response variable",
    xaxis = list(title = "Weight (1000 lb)"),
    yaxis = list(title = "Miles per gallon"),
    updatemenus = list(list(
      type = "dropdown",
      x = 1.05,
      y = 1,
      buttons = list(
        list(
          method = "update",
          args = list(list(y = list(mtcars$mpg)), 
                      list("yaxis.title.text" = "Miles per gallon")),
          label = "MPG"
        ),
        list(
          method = "update",
          args = list(list(y = list(mtcars$hp)), 
                      list("yaxis.title.text" = "Horsepower")),
          label = "Horsepower"
        )
      )
    ))
  )
```

### Heatmaps

This example shows the correlations between the four numeric iris measurements. Hovering over a cell reveals the exact correlation.

```{r}
iris_correlations <- iris |>
  select(where(is.numeric)) |>
  cor()

plot_ly(
  x = colnames(iris_correlations),
  y = rownames(iris_correlations),
  z = iris_correlations,
  type = "heatmap",
  colors = "Blues",
  zmin = -1,
  zmax = 1,
  hovertemplate = "%{y} vs %{x}: %{z:.2f}<extra></extra>"
) |>
  layout(
    title = "Correlation between iris measurements",
    xaxis = list(title = NULL),
    yaxis = list(title = NULL, autorange = "reversed")
  )
```

### Three-dimensional scatter plots

Explore three continuous dimensions with a 3D scatter plot. Use the mouse to
rotate the chart, then decide whether the additional dimension improves the
comparison over a 2D chart.

```{r}
plot_ly(
  iris,
  x = ~Sepal.Length,
  y = ~Sepal.Width,
  z = ~Petal.Length,
  color = ~Species,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 4, opacity = 0.8)
) |>
  layout(
    title = "Iris measurements in three dimensions",
    scene = list(
      xaxis = list(title = "Sepal length"),
      yaxis = list(title = "Sepal width"),
      zaxis = list(title = "Petal length")
    )
  )
```

### Point selection

Enable box or lasso selection with `dragmode`. The selected points stay
highlighted, which is useful for inspecting a subset before exporting or
linking it to another view in a Shiny application.

```{r}
plot_ly(
  mtcars,
  x = ~wt,
  y = ~mpg,
  text = ~model,
  type = "scatter",
  mode = "markers",
  marker = list(size = 10, opacity = 0.75),
  hovertemplate = "<b>%{text}</b><br>Weight: %{x:.2f}<br>MPG: %{y:.1f}<extra></extra>"
) |>
  layout(
    title = "Select vehicles with a box or lasso",
    dragmode = "select",
    xaxis = list(title = "Weight (1000 lb)"),
    yaxis = list(title = "Miles per gallon")
  )
```

### Challenge

1. Create a `Sepal.Width` vs `Sepal.Length` scatter plot of `iris`, colored by
   `Species`. Add an overall regression line by creating predictions with
   `lm()` and adding them with `add_lines()`.
2. Change the marker colors to a palette of your choice with `colors`.

::: {.callout-tip collapse="true"}

## Solution

```{r}
iris_model <- lm(Sepal.Width ~ Sepal.Length, data = iris)

iris_predictions <- tibble(
  Sepal.Length = seq(
    min(iris$Sepal.Length),
    max(iris$Sepal.Length),
    length.out = 100
  )
)
iris_predictions$Sepal.Width <- predict(iris_model, newdata = iris_predictions)

plot_ly(
  iris,
  x = ~Sepal.Length,
  y = ~Sepal.Width,
  color = ~Species,
  colors = c("#0072B2", "#D55E00", "#009E73"),
  type = "scatter",
  mode = "markers"
) |>
  add_lines(
    data = iris_predictions,
    x = ~Sepal.Length,
    y = ~Sepal.Width,
    name = "Overall linear fit",
    inherit = FALSE,
    line = list(color = "#333333")
  ) |>
  layout(
    title = "Iris sepal dimensions with an overall linear fit",
    xaxis = list(title = "Sepal length"),
    yaxis = list(title = "Sepal width")
  )
```

The points remain grouped by species, while the line summarizes the linear
relationship across all 150 flowers. Fitting separate lines by species would
answer a different question.

:::

## **highcharter**
### Basics

Create a scatter plot of the `iris` data using `hchart()`.

```{r}
iris |>
  hchart("scatter", hcaes(x = Sepal.Length, y = Sepal.Width, group = Species))
```

Create a bar chart from `mtcars` counting cars by number of cylinders.

```{r}
mtcars |>
  count(cyl) |>
  hchart("column", hcaes(x = as.character(cyl), y = n))
```

### Customization

Build the same cylinder bar chart using `highchart()` + `hc_add_series()`.

```{r}
count_tbl <- mtcars |>
  count(cyl)

highchart() |>
  hc_add_series(
    data = count_tbl$n,
    type = "column",
    name = "Cars"
  ) |>
  hc_xAxis(categories = as.character(count_tbl$cyl), title = list(text = "Cylinders")) |>
  hc_yAxis(title = list(text = "Count")) |>
  hc_title(text = "Cars by Cylinder Count")
```

Add custom tooltip formatting and a theme.

```{r}
highchart() |>
  hc_add_series(data = c(5, 8, 6, 9), type = "line", name = "Example") |>
  hc_xAxis(categories = c("Q1", "Q2", "Q3", "Q4")) |>
  hc_tooltip(pointFormat = "Value: <b>{point.y}</b>") |>
  hc_add_theme(hc_theme_flat())
```

### Grouped series

Use groups to make one series per cylinder count. `hchart()` turns the grouped
data into Highcharts series automatically.

```{r}
mtcars |>
  arrange(wt) |>
  hchart(
    "scatter",
    hcaes(x = wt, y = mpg, group = factor(cyl)),
    name = "Cars"
  ) |>
  hc_title(text = "Fuel economy by cylinder count") |>
  hc_xAxis(title = list(text = "Weight (1000 lb)")) |>
  hc_yAxis(title = list(text = "Miles per gallon")) |>
  hc_tooltip(
    pointFormat = "Weight: <b>{point.x:.2f}</b><br>MPG: <b>{point.y:.1f}</b>"
  )
```

### Time-series charts

Create multiple line series explicitly. A shared tooltip makes it easier to
compare the same date across both series, and x-axis zooming supports focused
exploration of a longer sequence.

```{r}
highchart() |>
  hc_add_series(
    economics_small,
    "line",
    hcaes(x = datetime_to_timestamp(date), y = unemployment),
    name = "Unemployed"
  ) |>
  hc_add_series(
    economics_small,
    "line",
    hcaes(x = datetime_to_timestamp(date), y = population),
    name = "Population"
  ) |>
  hc_xAxis(type = "datetime") |>
  hc_yAxis(title = list(text = "People (millions)")) |>
  hc_chart(zoomType = "x") |>
  hc_tooltip(shared = TRUE, valueDecimals = 1) |>
  hc_exporting(enabled = TRUE) |>
  hc_title(text = "Two time series on a common scale")
```

### Reference lines

Highlight a reference value with a plot line. This is useful when the reader
needs to compare data to a target, threshold, or policy limit.

```{r}
mtcars |>
  count(cyl) |>
  hchart("column", hcaes(x = as.character(cyl), y = n)) |>
  hc_yAxis(
    title = list(text = "Number of cars"),
    plotLines = list(list(
      value = 10,
      color = "#c0392b",
      width = 2,
      dashStyle = "ShortDash",
      label = list(text = "Reference: 10 cars")
    ))
  ) |>
  hc_title(text = "Cars by cylinder count")
```

### Data labels

Show values directly on a chart with data labels. Labels are most helpful when
there are only a few marks and reading an exact value is more important than
comparing bar heights.

```{r}
mtcars |>
  count(cyl) |>
  hchart("column", hcaes(x = as.character(cyl), y = n)) |>
  hc_title(text = "Cars by cylinder count") |>
  hc_xAxis(title = list(text = "Cylinders")) |>
  hc_yAxis(title = list(text = "Number of cars")) |>
  hc_plotOptions(column = list(dataLabels = list(enabled = TRUE))) |>
  hc_colors(c("#1565c0", "#ef6c00", "#2e7d32"))
```

### Challenge

Create a line chart of the average `mpg` for each `cyl` group and customize the tooltip to show one decimal place.

::: {.callout-tip collapse="true"}

## Solution
### Average fuel economy

```{r}
mtcars |>
  group_by(cyl) |>
  summarise(mean_mpg = mean(mpg), .groups = "drop") |>
  hchart("line", hcaes(x = as.character(cyl), y = mean_mpg)) |>
  hc_xAxis(title = list(text = "Cylinders")) |>
  hc_yAxis(title = list(text = "Average miles per gallon")) |>
  hc_tooltip(pointFormat = "Average MPG: <b>{point.y:.1f}</b>") |>
  hc_title(text = "Average fuel economy by cylinder count")
```

:::

## Comparison challenge

Build the same chart in both libraries:

- Data: `mtcars`
- Chart: `mpg` vs `wt`
- Color by: `cyl`
- Tooltip includes model name

::: {.callout-tip collapse="true"}

## Solution

### plotly

```{r}
plot_ly(
  mtcars,
  x = ~wt,
  y = ~mpg,
  color = ~factor(cyl),
  customdata = ~model,
  type = "scatter",
  mode = "markers"
) |>
  layout(
    title = "Fuel economy by vehicle weight",
    xaxis = list(title = "Weight (1000 lb)"),
    yaxis = list(title = "Miles per gallon"),
    legend = list(title = list(text = "Cylinders"))
  ) |>
  style(
    hovertemplate = paste0(
      "<b>%{customdata}</b><br>",
      "Weight: %{x:.2f}<br>",
      "MPG: %{y:.1f}<extra></extra>"
    )
  )
```

### highcharter

```{r}
mtcars |>
  hchart(
    "scatter",
    hcaes(x = wt, y = mpg, group = factor(cyl), name = model)
  ) |>
  hc_title(text = "Fuel economy by vehicle weight") |>
  hc_xAxis(title = list(text = "Weight (1000 lb)")) |>
  hc_yAxis(title = list(text = "Miles per gallon")) |>
  hc_tooltip(
    pointFormat = "<b>{point.name}</b><br>Weight: {point.x:.2f}<br>MPG: {point.y:.1f}"
  )
```

The two implementations use similar data mappings. Plotly makes it especially
easy to specify a custom hover template, while Highcharter closely follows the
Highcharts configuration API.

:::

Questions to answer:

1. Which code felt easier to write?
2. Which default output looked better to you?
3. Which package would you choose for your next report and why?

## Session

<details>
  <summary>Click here</summary>

```{r}
#| label: session

sessionInfo()
```

</details>

