---
title: "My report"
description: "The Day-1 document, now a page on the site."
---

```{r}
#| label: setup
#| message: false
library(dplyr)
library(ggplot2)
library(ggokabeito)   # color-blind-safe (Okabe-Ito) scale
data(penguins)
penguins <- penguins |> filter(!is.na(bill_len), !is.na(bill_dep), !is.na(body_mass))
```

This page stands in for the report you built on Day 1. Copy your own `my-report.qmd` over this file
and re-render: the document you wrote as a standalone PDF becomes one page of a website, with no
change to its content.

Only the YAML header needs to change. A page inside a project inherits `format:` from `_quarto.yml`,
so a `format: typst` left over from the Day-1 document would keep building a PDF instead of a page.

```{r}
#| label: fig-bill
#| fig-cap: "Bill depth against bill length, by species."
#| fig-alt: "Scatter plot of bill depth versus bill length in millimeters, colored by species. The three species form separate clusters."
ggplot(penguins, aes(bill_len, bill_dep, color = species)) +
  geom_point(alpha = 0.8) +
  scale_color_okabe_ito() +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)", color = NULL)
```

@fig-bill still resolves, because the reference and its figure share one page.
