---
title: "data.table"
author: "Markus Mayrhofer"
description: "Fast, memory-efficient tabular data wrangling in R, compared with base R and the tidyverse."
format: revealjs
css: assets/custom.css
execute:
  echo: true
  eval: false
---

## Data {visibility="hidden"}

```{r}
#| label: slide-data
#| eval: true
#| echo: false
#| message: false
library(data.table)

simulate_genotyping <- function(n_samples = 500, n_variants = 4000, seed = 42) {
  set.seed(seed)
  pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")
  samples <- data.table(
    sample_id  = sprintf("S%05d", seq_len(n_samples)),
    population = sample(pops, n_samples, replace = TRUE,
                        prob = c(0.22, 0.14, 0.20, 0.30, 0.14)),
    sex        = sample(c("M", "F"), n_samples, replace = TRUE),
    batch      = sample(sprintf("batch%02d", 1:8), n_samples, replace = TRUE),
    phenotype  = rbinom(n_samples, 1, 0.35)
  )
  chroms <- c(1:22, "X")
  conseq <- c("synonymous", "missense", "stop_gained", "splice_region", "intron", "UTR")
  variants <- data.table(
    variant_id  = sprintf("rs%07d", seq_len(n_variants)),
    chrom       = sample(chroms, n_variants, replace = TRUE),
    pos         = sample(1e6:2.5e8, n_variants, replace = TRUE),
    ref         = sample(c("A", "C", "G", "T"), n_variants, replace = TRUE),
    alt         = sample(c("A", "C", "G", "T"), n_variants, replace = TRUE),
    gene        = sprintf("GENE%04d", sample(1:1500, n_variants, replace = TRUE)),
    consequence = sample(conseq, n_variants, replace = TRUE,
                         prob = c(0.30, 0.25, 0.03, 0.07, 0.25, 0.10)),
    maf         = round(rbeta(n_variants, 0.6, 6), 4)
  )
  flip <- variants$alt == variants$ref
  variants[flip, alt := fifelse(ref == "A", "G", "A")]
  geno <- CJ(sample_id = samples$sample_id, variant_id = variants$variant_id)
  geno[variants, maf := i.maf, on = "variant_id"]
  n <- nrow(geno)
  p <- geno$maf
  u <- runif(n)
  geno[, genotype := fifelse(u < (1 - p)^2, 0L,
                       fifelse(u < (1 - p)^2 + 2 * p * (1 - p), 1L, 2L))]
  geno[runif(n) < 0.02, genotype := -1L]
  geno[, read_depth       := rpois(n, lambda = 30)]
  geno[, genotype_quality := pmin(99L, rpois(n, lambda = 45))]
  geno[, maf := NULL]
  setcolorder(geno, c("sample_id", "variant_id", "genotype", "read_depth", "genotype_quality"))
  list(genotypes = geno[], variants = variants[], samples = samples[])
}

dat       <- simulate_genotyping(n_samples = 500, n_variants = 4000, seed = 42)
samples   <- dat$samples
variants  <- dat$variants
genotypes <- dat$genotypes
```

## This session {.smaller}

`data.table` is an R package for working with **rectangular / tabular data**.

::: {.incremental}
1. Demonstrate the core table operations: **choose rows, choose columns, add columns, group, summarise, join**.
2. Compare the **same task** in three styles: **base R**, **tidyverse**, and **data.table**.
3. See what happens under the hood: **copies, memory, indexing, keys**.
4. Simulate large biological data where the differences are visible.
:::

. . .

Work through the companion lab, which contains the executable examples,
benchmarks, and exercises.

::: {.aside}
`data.table` may run before the tidyverse session, but the tidyverse code is
shown as a *reference* since many of you will already be familiar with it.
:::

## Where data.table comes from {.smaller}

- Created by **Matt Dowle**; version 1.0 on CRAN in **April 2006**. Long co-developed with **Arun Srinivasan**, now maintained by a wider team.
- Born from a practical problem: base R's `data.frame` **copies** data on many operations, which is slow and memory-hungry on large tables.
- Written largely in **C**, **multi-threaded**, changes columns **by reference** (no copy), and ships fast I/O (`fread` / `fwrite`).
- Deliberately **few dependencies** (just base R), part of why it stays fast and stable.

. . .

`data.table` has a terse `DT[i, j, by]` syntax, which is what we will demonstrate.

## A table is just rows, columns, groups, joins {.smaller}

Rows are observations, columns are variables, a cell is one value. Almost every
data-wrangling question is one of four actions:

| Question | Operation |
|---|---|
| Which rows do I keep? | **filter rows** |
| Which columns do I keep or compute? | **select / mutate columns** |
| Should I calculate separately for each group? | **group by** |
| Do I need information from another table? | **join** |

Today we express each of these three ways and compare them.

## A bioinformatics data example. {.smaller}

Genetic data produces **large tables** frequently. We'll simulate a data set.

| Term | Meaning in this session |
|---|---|
| sample | one person or specimen |
| variant | one genomic position where samples may differ |
| genotype | alternate copies: 0, 1, or 2; **-1 = missing** |
| read depth | how many sequencing reads cover a site |
| genotype quality | confidence score for the call |
| allele frequency | average alternate-copy frequency across samples |

## Three linked tables {.smaller .scrollable}

- **samples**: one row per sample (population, sex, batch, case/control)
- **variants**: one row per variant (chrom, position, gene, consequence, frequency)
- **genotypes**: one row per sample x variant (genotype, read depth, quality) -- the **tall** table

```r
dat       <- simulate_genotyping(n_samples = 500, n_variants = 4000)
samples   <- dat$samples
variants  <- dat$variants
genotypes <- dat$genotypes      # 500 x 4000 = 2,000,000 rows
```

::: {.panel-tabset}

### samples

```{r}
#| eval: true
#| echo: false
head(samples, 50)
```

### variants

```{r}
#| eval: true
#| echo: false
head(variants, 50)
```

### genotypes

```{r}
#| eval: true
#| echo: false
head(genotypes, 50)
```

:::

::: {.aside}
The generator is provided with the lab: reproducible, no downloads.
:::




## A table is just rows, columns, groups, joins {.smaller}

Rows are observations, columns are variables, a cell is one value. Almost every
data-wrangling question is one of four actions:

| Question | Operation |
|---|---|
| Which rows do I keep? | **filter rows** |
| Which columns do I keep or compute? | **select / mutate columns** |
| Should I calculate separately for each group? | **group by** |
| Do I need information from another table? | **join** |

Today we express each of these three ways and compare them.




## `data.table` syntax: `DT[i, j, by]`

```r
DT[ i ,  j ,  by ]
|  |    |    `-- group by these columns
|  |    `------- compute, select, or update columns
|  `------------ keep or match these rows
`--------------- the data.table object
```

Read it as:

> Take rows **`i`**, do **`j`**, separately within each **`by`** group.

No grouping? Leave `by` out. No row filter? Leave `i` empty but keep the comma: `DT[, j]`.

. . .

`.N` = rows in the current group. `.()` = shorthand for `list()`.

## Filter rows {.smaller}

**Task:** keep called homozygous-alt rows with adequate depth.

```r
# base R
genotypes_df[genotypes_df$genotype == 2 & genotypes_df$read_depth >= 20, ]

# tidyverse
genotypes_df |> filter(genotype == 2, read_depth >= 20)

# data.table -- the row condition goes in i
genotypes[genotype == 2 & read_depth >= 20]
```

Under the hood: all three build a TRUE/FALSE vector over the rows and keep the
`TRUE` ones. All three return a **new** table -- the original is untouched -- so
you pay memory for the rows you keep. You typically assign the returned table a name.


## Select columns {.smaller}

**Task:** keep only a few of the variant columns.

```r
# select a few columns
variants_df[, c("variant_id", "gene", "maf")]      # base R
variants_df |> select(variant_id, gene, maf)       # tidyverse
variants[, .(variant_id, gene, maf)]               # data.table
```

Under the hood: all three return a **new** table object -- but data vectors 
are not duplicated until necessary.


## Create columns {.smaller}

```r
# add a column

# base R: adds the column, rebinds the name
variants_df$is_rare <- variants_df$maf < 0.01               

# tidyverse: you must assign/name the result
variants_df <- variants_df |> mutate(is_rare = maf < 0.01)  

# data.table: modifies in place, no assignment
variants[, is_rare := maf < 0.01]                           
```

. . .

| Style | Adding a column |
|---|---|
| base R | shallow copy: columns are shared until modified, then only that column is duplicated |
| tidyverse | same, but returns a new object -- you must assign it |
| data.table `:=` | no new object at all; every reference sees the change |

`:=` needs no assignment and changes the original in place. Convenient, but every
other reference to that table sees the change. Use `copy(variants)` to avoid.

## Grouped summaries {.smaller}

**Task:** per sample, mean read depth among called genotypes.

```r
# base R
called <- genotypes_df[genotypes_df$genotype >= 0, ]
aggregate(read_depth ~ sample_id, data = called, FUN = mean)

# tidyverse
genotypes_df |> filter(genotype >= 0) |>
  group_by(sample_id) |> summarise(mean_depth = mean(read_depth))

# data.table -- i filters, j computes, by groups
genotypes[genotype >= 0, .(mean_depth = mean(read_depth)), by = sample_id]
```

Split by group, compute within each group, combine. `data.table` does this
with low overhead, especially when the data are already keyed.

## Joins: add columns from another table {.smaller}

```r
# base R
merge(genotypes_df, variants_df[, c("variant_id","consequence","gene")], 
        by = "variant_id")

# tidyverse
genotypes_df |> inner_join(select(variants_df, variant_id, consequence, gene), 
        by = "variant_id")

# data.table -- x[y, on = "key"]
variants[genotypes, on = "variant_id",
         .(sample_id, variant_id, genotype, consequence, gene)]
```


## Keys: sort once, look up fast {.smaller}

A **key** is one or more columns a `data.table` is physically **sorted by**.

```r
setkey(genotypes, sample_id)   # reorders genotypes in place, by reference
key(genotypes)                 # "sample_id"
```

Because the rows are sorted, `data.table` can use **binary search** and
**sorted-merge joins** instead of scanning every row:

```r
genotypes["S00001"]            # fast lookup of one sample's rows
setkey(variants, variant_id)   # then a keyed join is a fast sorted merge
```

. . .

A key speeds up **lookups**, **repeated joins**, and **grouping** on the key
column. You do not always need one (`on =` joins ad-hoc), but set a key when you
will hit the same column many times, so the sort is paid once.

## Benchmarks: time *and* memory {.smaller}

`bench::mark()` records **elapsed time**, **memory allocated**, and **garbage collections**.

```r
bench::mark(
  base_R     = { ... tapply(...) ... },
  tidyverse  = genotypes_df |> filter(genotype >= 0) |>
                 group_by(variant_id) |> summarise(af = sum(genotype)/(2*n())),
  data.table = genotypes[genotype >= 0, .(af = sum(genotype)/(2*.N)), by = variant_id],
  check = FALSE, iterations = 3
)
```

- **Time** matters because slow code interrupts your work.
- **Memory** matters because large tables can *fail* when R runs out of RAM.

## What to expect {.smaller}

::: {.incremental}
- On **a few thousand rows**, all three are fast enough: pick your preference.
- The gap grows with **row count**, **repeated joins**, **grouped calculations**, and **in-place updates**.
- `data.table` often wins on time and/or memory at scale, and `fread`/`fwrite` are the most reliable everyday win.
:::


## Takeaways

::: {.incremental}
- Every table operation is about **rows, columns, groups, or joins**.
- `data.table` packs them into `DT[i, j, by]`.
- `:=` modifies **by reference**: saves memory, changes the original.
- **Keys** sort/index tables for fast repeated joins and lookups.
:::

. . .

**Next:** the lab, then `vignette("datatable-intro")` and `?data.table`.

## {background-image="/assets/images/network.svg"}

::: {.v-center .center}

[Thank you!]{.largest}

[Questions?]{.larger}

[{{< meta current_year >}} • [SciLifeLab](https://www.scilifelab.se/) • [NBIS](https://nbis.se/) • [RaukR](https://nbisweden.github.io/raukr-2026)]{.smaller}

:::
